---
title: "Labour Market Indicators for Youth"
subtitle: "Automated Statistics Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
params:
  data: null
  descriptives: null
  tables: null
  unemployment_plot: null
  employment_plot: null

---

# Overview

This report presents key labour market indicators for youth (ages 16–24) across Swedish municipalities.  
All results are generated through a fully reproducible `targets` pipeline.

The analysis includes:

- population structure  
- unemployment and employment rates  
- breakdowns by sex and age group  
- descriptive statistics  
- tables and visualisations  

---

# Data Summary

```{r}
params$data |>
  dplyr::glimpse()
```

---

# Descriptive Statistics

```{r}
params$descriptives
```

---

# Tables

```{r}
htmltools::HTML(params$tables$main_table_html)
```

---

# Plots

Distribution of Municipal Youth Unemployment Rates


![](`r params$unemployment_plot`)

Distribution of Municipal Youth Employment Rates by Year


![](`r params$employment_plot`)

---

# Methodology

This report is generated using:

- **R** (tidyverse, gt, ggplot2)
- **targets** for reproducible workflows
- **Quarto** for reporting
The dataset is synthetic and constructed to mimic municipal‑level labour market indicators.

---

# Reporducibility

All results in this report are produced automatically through the reproducible
pipeline defined in `_targets.R`, using the `{targets}` package.

The report itself is rendered using precomputed results passed via `execute_params`.

---