```markdown
---
title: "Fermentation KPI Analysis"
author: "Ioannis"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)

# Load data
data <- read_csv("../../data/simulated_fermentation_data.csv")
```

# 📘 Overview
This report summarizes key fermentation KPIs and visualizes process trends using simulated fed‑batch data.  
It demonstrates a reproducible analytics workflow suitable for bioprocess development, MSAT, and GMP digitalization environments.

---

# 🧪 Methods
The analysis includes:
- Importing and cleaning simulated fermentation data  
- Calculating essential KPIs (biomass, glucose, DO, pH)  
- Visualizing process behavior over time  
- Interpreting trends from a bioprocess engineering perspective  

All analysis is performed in **R** using tidyverse and ggplot2.

---

# 📊 Data Overview

```{r}
head(data)
summary(data)
```

The dataset includes:
- **Biomass (g/L)**  
- **Glucose (g/L)**  
- **Dissolved Oxygen (%)**  
- **pH**  
- **Temperature (°C)**  
- **Time (hr)**  

---

# 📈 KPI Summary

```{r}
kpis <- data %>%
  summarise(
    max_biomass = max(biomass_gL),
    min_glucose = min(glucose_gL),
    avg_do = mean(do_pct),
    final_pH = last(ph)
  )

kpis
```

---

# 📈 Biomass Growth Curve

```{r, echo=FALSE}
ggplot(data, aes(time_hr, biomass_gL)) +
  geom_line(color = "steelblue", size = 1.2) +
  theme_minimal() +
  labs(title = "Biomass Growth Over Time",
       x = "Time (hr)",
       y = "Biomass (g/L)")
```

---

# 📉 Glucose Consumption

```{r, echo=FALSE}
ggplot(data, aes(time_hr, glucose_gL)) +
  geom_line(color = "firebrick", size = 1.2) +
  theme_minimal() +
  labs(title = "Glucose Consumption Over Time",
       x = "Time (hr)",
       y = "Glucose (g/L)")
```

---

# Dissolved Oxygen Profile

```{r, echo=FALSE}
ggplot(data, aes(time_hr, do_pct)) +
  geom_line(color = "darkgreen", size = 1.2) +
  theme_minimal() +
  labs(title = "Dissolved Oxygen Profile",
       x = "Time (hr)",
       y = "DO (%)")
```

---

# 📊 Combined Process Overview

```{r, echo=FALSE}
data_long <- data %>%
  pivot_longer(cols = c(biomass_gL, glucose_gL, do_pct),
               names_to = "variable",
               values_to = "value")

ggplot(data_long, aes(time_hr, value, color = variable)) +
  geom_line(size = 1.1) +
  theme_minimal() +
  labs(title = "Overall Fermentation Profile",
       x = "Time (hr)",
       y = "Value",
       color = "Variable")
```

---

# 🧠 Discussion
The fermentation shows expected fed‑batch behavior:

- **Biomass** increases steadily, indicating healthy growth.  
- **Glucose** decreases as biomass consumes substrate.  
- **DO** declines as oxygen demand rises with cell density.  
- **pH** shows a slight downward trend, consistent with metabolic acidification.  

These KPIs provide a quick snapshot of process performance and can support routine monitoring, troubleshooting, and comparison across batches.

---

# 🏁 Conclusion
This report demonstrates a clean, reproducible workflow for analyzing fermentation data using R.  
The approach can be extended to:

- Batch‑to‑batch comparison  
- Anomaly detection  
- Soft‑sensor development  
- Automated dashboards (Shiny)  
- SPC charting for GMP environments  

This project highlights the integration of **bioprocess engineering expertise** with **modern data analytics**, aligning with digitalization and MSAT expectations.
