---
title: "Retail Sales Analytics – SQL + R (PostgreSQL)"
author: "Ioannis Dogaris"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(dbplyr)
library(here)
```

## 1. Project setup

### 1.1 Connect to PostgreSQL

```{r connection}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "retail_analytics",
  host   = "localhost",
  port   = 5432,
  user   = "postgres",
  password = "1234"
)
con
```

### 1.2 Define SQL helper function

```{r sql-helper}
data_dir <- here::here("04-retail-analytics-sql/data")

run_sql_file <- function(path) {
  sql <- readr::read_file(path)

  # Split only on semicolons that end a line
  statements <- unlist(strsplit(sql, ";(?=(?:[^']|'[^']*')*$)", perl = TRUE))

  # Clean whitespace
  statements <- trimws(statements)

  # Remove empty statements
  statements <- statements[nchar(statements) > 0]

  # Execute each statement
  purrr::walk(statements, ~ dbExecute(con, .x))
}
```

### 1.3 Load CSV data

```{r load-data}
data_dir <- here::here("04-retail-analytics-sql/data")

products    <- readr::read_csv(file.path(data_dir, "products.csv"), show_col_types = FALSE)
customers   <- readr::read_csv(file.path(data_dir, "customers.csv"), show_col_types = FALSE)
orders      <- readr::read_csv(file.path(data_dir, "orders.csv"), show_col_types = FALSE)
order_items <- readr::read_csv(file.path(data_dir, "order_items.csv"), show_col_types = FALSE)
```

### 1.4 Apply schema and insert data

```{r schema-sql}
sql_dir <- here::here("04-retail-analytics-sql/sql")

# Run SQL scripts (they execute directly, no need to store return values)
run_sql_file(file.path(sql_dir, "01_schema.sql"))
run_sql_file(file.path(sql_dir, "02_data_cleaning.sql"))
run_sql_file(file.path(sql_dir, "03_joins_aggregations.sql"))
run_sql_file(file.path(sql_dir, "04_window_functions.sql"))

# Load data into tables
dbWriteTable(con, "products", products, append = TRUE, row.names = FALSE)
dbWriteTable(con, "customers", customers, append = TRUE, row.names = FALSE)
dbWriteTable(con, "orders", orders, append = TRUE, row.names = FALSE)
dbWriteTable(con, "order_items", order_items, append = TRUE, row.names = FALSE)
```


## 2. Exploratory data analysis

### 2.1 Basic summaries

```{r eda-summary}
tbl(con, "products") %>% glimpse()
tbl(con, "customers") %>% glimpse()
tbl(con, "orders") %>% glimpse()
tbl(con, "order_items") %>% glimpse()
```

```{r eda-counts}
tbl(con, "orders") %>%
  count(order_region, order_channel) %>%
  arrange(desc(n)) %>%
  collect()
```

### 2.2 Missing values

```{r missing-values}
check_missing <- function(tbl_name) {
  df <- tbl(con, tbl_name) %>% collect()
  tibble(
    table = tbl_name,
    column = names(df),
    missing = purrr::map_int(df, ~ sum(is.na(.x))),
    pct_missing = round(missing / nrow(df) * 100, 1)
  )
}

bind_rows(
  check_missing("products"),
  check_missing("customers"),
  check_missing("orders"),
  check_missing("order_items")
)
```

### 2.3 Distribution of basket value

```{r distribution-plots, fig.height=4, fig.width=7}
sales_fact <- tbl(con, "v_sales_fact")

basket <- sales_fact %>%
  group_by(order_id) %>%
  summarise(basket_value = sum(line_total), .groups = "drop") %>%
  collect()

ggplot(basket, aes(x = basket_value)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "white") +
  labs(
    title = "Distribution of Basket Value",
    x = "Basket value",
    y = "Count of orders"
  ) +
  theme_minimal()
```

## 3. SQL + R integration

### 3.1 Raw SQL

```{r sql-direct}
top_products_sql <- "
SELECT
  product_name,
  category,
  SUM(line_total) AS total_revenue
FROM v_sales_fact
GROUP BY product_name, category
ORDER BY total_revenue DESC
LIMIT 5;
"

top_products <- dbGetQuery(con, top_products_sql)
top_products
```

### 3.2 dbplyr

```{r dbplyr}
sales_tbl <- tbl(con, "v_sales_fact")

revenue_by_region <- sales_tbl %>%
  group_by(order_region) %>%
  summarise(
    total_revenue = sum(line_total),
    orders = n_distinct(order_id),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue)) %>%
  collect()

revenue_by_region
```

## 4. Visualizations

### 4.1 Revenue over time

```{r revenue-over-time, fig.height=4, fig.width=7}
daily_rev <- tbl(con, "v_daily_revenue") %>% collect()

ggplot(daily_rev, aes(x = order_date, y = daily_revenue)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(color = "darkgreen") +
  labs(
    title = "Daily Revenue",
    x = "Date",
    y = "Revenue"
  ) +
  theme_minimal()
```

### 4.2 Top products bar chart

```{r top-products-plot, fig.height=4, fig.width=7}
top_products_plot <- top_products %>%
  mutate(product_name = fct_reorder(product_name, total_revenue))

ggplot(top_products_plot, aes(x = product_name, y = total_revenue, fill = category)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Products by Revenue",
    x = "Product",
    y = "Total revenue"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

### 4.3 Customer segments

```{r customer-segments, fig.height=4, fig.width=7}
customer_seg_sql <- "
WITH customer_spend AS (
  SELECT
    customer_id,
    first_name,
    last_name,
    SUM(line_total) AS total_revenue
  FROM v_sales_fact
  GROUP BY customer_id, first_name, last_name
)
SELECT
  customer_id,
  first_name,
  last_name,
  total_revenue,
  CASE
    WHEN total_revenue >= 60 THEN 'High'
    WHEN total_revenue >= 30 THEN 'Medium'
    ELSE 'Low'
  END AS segment
FROM customer_spend;
"

customer_segments <- dbGetQuery(con, customer_seg_sql)

ggplot(customer_segments, aes(x = segment, y = total_revenue, fill = segment)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Customer Revenue by Segment",
    x = "Segment",
    y = "Total revenue"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

### 4.4 Heatmap of sales by weekday and hour

```{r heatmap-weekday-hour, fig.height=5, fig.width=7}
heatmap_sql <- "
SELECT
  TO_CHAR(order_date, 'D') AS weekday_num,
  EXTRACT(HOUR FROM order_time) AS hour_of_day,
  SUM(line_total) AS total_revenue
FROM v_sales_fact
GROUP BY TO_CHAR(order_date, 'D'), EXTRACT(HOUR FROM order_time);
"

heatmap_df <- dbGetQuery(con, heatmap_sql) %>%
  mutate(
    weekday_num = as.integer(weekday_num),
    hour_of_day = as.integer(hour_of_day),
    weekday = factor(
      weekday_num,
      levels = 1:7,
      labels = c('Sun','Mon','Tue','Wed','Thu','Fri','Sat')
    )
  )

ggplot(heatmap_df, aes(x = hour_of_day, y = weekday, fill = total_revenue)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Revenue Heatmap by Weekday and Hour",
    x = "Hour of day",
    y = "Weekday",
    fill = "Revenue"
  ) +
  theme_minimal()
```

## 5. Insights and recommendations

```{r insights, echo=FALSE}
top_products
revenue_by_region
customer_segments %>% count(segment)
```

- Coffee and snack products dominate revenue.
- Online and North-region customers contribute strongly.
- A small group of customers forms a high-value segment.
- Basket values are higher when coffee products are included.
- Revenue peaks in morning and early evening hours.

```{r disconnect, echo=FALSE}
dbDisconnect(con)
```
